---
title: "Canid Diet Experiment NOG/COG Results"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(echo = F, include = T, message = F, warning = F, cache = T)
```

```{r libraries}
library(tximport)
library(ggplot2)
library(ggrepel)
library(readr)
library(DESeq2)
library(dplyr)
library(vegan)
library(kableExtra)
library(ggbeeswarm)
```

```{r tximport_og}
txi_og <- read_tsv("outputs/tx2gene/eggnog_tx2gene_og.tsv")
txi_og <- filter(txi_og, grepl("@", GENEID))
canids <- read_csv("inputs/tximport_samples/canid_samples.csv")
canid_counts_og <- tximport(files = canids$filepath, type = "salmon", tx2gene = txi_og)
```

```{r export_counts}
canid_counts_raw <- canid_counts_og$counts
colnames(canid_counts_raw) <- canids$sample
canid_counts_raw <- as.data.frame(canid_counts_raw)
canid_counts_raw$ortholog <- rownames(canid_counts_raw)
write_tsv(canid_counts_raw, "outputs/ortholog_counts/canid_og_counts_raw.tsv")
```

```{r diff_abund_og}
ddsTxi_og <- DESeqDataSetFromTximport(canid_counts_og,
                                      colData = canids,
                                      design = ~ species + diet)
dds_og <- DESeq(ddsTxi_og)
# resultsNames(dds_og) # lists the coefficients
```

```{r norm_counts}
dds_og <- estimateSizeFactors(dds_og)
dds_og_counts <- counts(dds_og, normalized=TRUE)
colnames(dds_og_counts) <- colData(dds_og)$sample
dds_og_counts <- as.data.frame(dds_og_counts)
dds_og_counts$OG <- rownames(dds_og_counts)
write_tsv(dds_og_counts, "outputs/ortholog_counts/canid_og_counts_norm.tsv")
```

```{r pca_og}
vsd <- vst(dds_og)
pca_data <- plotPCA(vsd, intgroup = c("species", "diet", "Individual"), returnData=TRUE)
percentVar <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(PC1, PC2, shape=species, color=diet)) + 
  theme_minimal() +
  geom_point(size=3) +
  geom_text_repel(aes(label = Individual)) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) +
  ggtitle("PCA plot of NOG/COG orthologs in canid diet study")
```

Permanova using `species + diet` as a model shows:
```{r permanova}
vsd_assay <- t(assay(vsd))
rownames(vsd_assay) <- colData(vsd)$sample
veg_dist <- vegdist(vsd_assay, "bray")

canids <- canids[match(rownames(vsd_assay), canids$sample), ] # sort info by colnames
canids$species <- as.factor(canids$species)              # set factors for model
canids$diet <- as.factor(canids$diet)
canids$day <- as.factor(canids$day )

perm <- adonis(veg_dist ~ species + diet, 
               data = canids, 
               permutations = 10000)
write.table(perm$aov.tab, "outputs/permanova/canid_og_permanova.tsv", quote = F)
kable(perm$aov.tab) %>% kable_styling()
```


### Diet: Wild vs. Chow -- Differential abundance

2,828,230 genes annotated to 185,870 NOG/COG orthologs. 3,014 were significantly 
differentially abundant between wild and chow diet, controlling for species. 
475 were more 
abundant in wild diet, while 2539 were more abundant in the chow diet.

```{r og_diet_res}
res <- results(dds_og, name="diet_Wild_vs_Chow")
res <- res[order(res$pvalue), ]

res_sig <- res %>%
  as.data.frame %>%
  mutate(ortholog = rownames(.)) %>%
  filter(grepl("@", ortholog)) %>%
  filter(padj < .05) %>%
  mutate(id = seq(1:nrow(.)))

res_og <- res %>%
  as.data.frame %>%
  mutate(ortholog = rownames(.)) %>%
  filter(grepl("@", ortholog))
```

### Species: Wolf vs. Dog -- Differential abundance

2349 genes were significantly differentially abundant between wolf and dog, 
blocking for diet. 1485 were more abundant in wolf, while 864 were more
abundant in dogs.

```{r og_species_res}
res_species <- results(dds_og, name="species_Wolf_vs_Dog")
res_species <- res_species[order(res_species$pvalue), ]

res_species_sig <- res_species %>%
  as.data.frame %>%
  mutate(ortholog = rownames(.)) %>%
  filter(grepl("@", ortholog)) %>%
  filter(padj < .05) %>%
  mutate(id = seq(1:nrow(.)))

res_species_og <- res_species %>%
  as.data.frame %>%
  mutate(ortholog = rownames(.)) %>%
  filter(grepl("@", ortholog))

#table(res_species_sig$log2FoldChange < 0) 
```

```{r write_tsvs}
write_tsv(res_sig, "outputs/diff_abund/canid/canid_diet_og_sig.tsv")
write_tsv(res_og, "outputs/diff_abund/canid/canid_diet_og_all.tsv")
write_tsv(res_species_sig, "outputs/diff_abund/canid/canid_species_og_sig.tsv")
write_tsv(res_species_og, "outputs/diff_abund/canid/canid_species_og_all.tsv")
```

### Do the species change in reaction to their diets in different amounts?

Instead of comparing across species or across diet, we can also ask what types of
changes each species encountered when switching diets, and which changed the most.
To do this, I defined a new variable, "interaction", which will allow me to 
compare dog control vs. dog on switched diet, and wolf control vs. wolf on a 
switched diet. This can't be built into the first model that we used above, 
as that model would not be full rank. Dog has many more orthologs that are 
significantly differentially abundanct,

```{r differential_abundance2, include = F}
ddsTxi2_og <- DESeqDataSetFromTximport(canid_counts_og,
                                   colData = canids,
                                   design = ~ interaction)
dds2_og <- DESeq(ddsTxi2_og)
```


```{r}
# resultsNames(dds2_og) # lists the coefficients

dog <-  results(dds2_og, name="interaction_dog_switch_vs_dog_ctrl")
dog_sig <- subset(dog, padj < .05)
dog$animal <- "dog"
dog_sig$animal <- "dog"
wolf <- results(dds2_og, contrast=c("interaction", "wolf_switch", "wolf_ctrl"))
wolf_sig <- subset(wolf, padj < .05)
wolf_sig$animal <- "wolf"
wolf$animal <- "wolf"
dog_wolf <- rbind(dog, wolf)
dog_wolf_sig <- rbind(dog_sig, wolf_sig)
dog_wolf <- as.data.frame(dog_wolf)
dog_wolf_sig <- as.data.frame(dog_wolf_sig)

ggplot(data = dog_wolf, aes(x = animal, y = log2FoldChange, color = abs(padj))) +
  geom_quasirandom(aes(color = ifelse(abs(padj) < .05, "significant", "non-significant")), alpha = .1) +
  labs(y = "Species", x = "log2 Fold Change", 
       title = "Distribution of log2 Fold Change for significant genes", 
       subtitle = "Dog wild diet vs. dog chow diet, and wolf chow diet vs. wolf wild diet") +
  theme_classic() +
  theme(plot.title.position = "plot") +
  scale_color_viridis_d() + 
  guides(color=guide_legend(title="Significance"))
```

```{r}
tally_dog_wolf_sig<- dog_wolf_sig %>%
  mutate(direction = ifelse(log2FoldChange < 0, "down", "up")) %>%
  group_by(animal, direction) %>%
  tally()

kable(tally_dog_wolf_sig) %>% kable_styling()
```