---
title: "Mouse Diet Experiment EggNog Results"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(echo = F, include = T, message = F, warning = F, cache = T)
```

```{r libraries}
library(tximport)
library(ggplot2)
library(ggrepel)
library(readr)
library(DESeq2)
library(dplyr)
library(vegan)
library(kableExtra)
```

```{r tximport_og}
txi_og <- read_tsv("outputs/tx2gene/eggnog_tx2gene_og.tsv")
txi_og <- filter(txi_og, grepl("@", GENEID))
mouse <- read_csv("inputs/tximport_samples/mouse_samples.csv")
mouse_counts_og <- tximport(files = mouse$filepath, type = "salmon", tx2gene = txi_og)
```

```{r export_counts}
mouse_counts_raw <- mouse_counts_og$counts
colnames(mouse_counts_raw) <- mouse$SampleID
mouse_counts_raw <- as.data.frame(mouse_counts_raw)
mouse_counts_raw$ortholog <- rownames(mouse_counts_raw)
write_tsv(mouse_counts_raw, "outputs/ortholog_counts/mouse_og_counts_raw.tsv")
```

```{r diff_abund_og}
ddsTxi_og <- DESeqDataSetFromTximport(mouse_counts_og,
                                      colData = mouse,
                                      design = ~ species + diet)
dds_og <- DESeq(ddsTxi_og)
# resultsNames(dds_og) # lists the coefficients
```

```{r norm_counts}
dds_og <- estimateSizeFactors(dds_og)
dds_og_counts <- counts(dds_og, normalized=TRUE)
colnames(dds_og_counts) <- colData(dds_og)$sample
dds_og_counts <- as.data.frame(dds_og_counts)
dds_og_counts$OG <- rownames(dds_og_counts)
write_tsv(dds_og_counts, "outputs/ortholog_counts/mouse_og_counts_norm.tsv")
```

```{r pca_og}
vsd <- vst(dds_og)
pca_data <- plotPCA(vsd, intgroup = c("species", "actual_diet", "individual"), returnData=TRUE)
percentVar <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(PC1, PC2, shape=species, color=actual_diet)) + 
  theme_minimal() +
  geom_point(size=3) +
  geom_text_repel(aes(label = individual)) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) +
  ggtitle("PCA plot of NOG/COG orthologs in mouse diet study")
```

```{r permanova}
vsd_assay <- t(assay(vsd))
rownames(vsd_assay) <- colData(vsd)$SampleID
veg_dist <- vegdist(vsd_assay, "bray")

mouse <- mouse[match(rownames(vsd_assay), mouse$SampleID), ] # sort info by colnames
mouse$species <- as.factor(mouse$species)              # set factors for model
mouse$actual_diet <- as.factor(mouse$actual_diet)
mouse$species <- as.factor(mouse$species)

perm <- adonis(veg_dist ~ species + actual_diet, 
               data = mouse, 
               permutations = 10000)
write.table(perm$aov.tab, "outputs/permanova/mouse_og_permanova.tsv", quote = F)
kable(perm$aov.tab) %>% kable_styling()
```

## Diet: Wild vs. Chow -- Differential abundance

```{r og_diet_res}
res <- results(dds_og, name="diet_Wild_vs_Chow")
res <- res[order(res$pvalue), ]

res_sig <- res %>%
  as.data.frame %>%
  mutate(ortholog = rownames(.)) %>%
  filter(grepl("@", ortholog)) %>%
  filter(padj < .05) 

res_og <- res %>%
  as.data.frame %>%
  mutate(ortholog = rownames(.)) %>%
  filter(grepl("@", ortholog))
```

`r nrow(txi_og)` genes annotated to `r length(unique(txi_og$GENEID))` NOG/COG orthologs. `r nrow(res_sig)` were significantly 
differentially abundant between wild and chow diet, controlling for species. 
 `r table(res_sig$log2FoldChange > 0)[2]` were more 
abundant in wild diet, while  `r table(res_sig$log2FoldChange > 0)[1]` were more abundant in the chow diet.

## Species: Wildmouse vs. labmouse -- Differential abundance



```{r og_species_res}
res_species <- results(dds_og, name="species_wildmouse_vs_labmouse")
res_species <- res_species[order(res_species$pvalue), ]

res_species_sig <- res_species %>%
  as.data.frame %>%
  mutate(ortholog = rownames(.)) %>%
  filter(grepl("@", ortholog)) %>%
  filter(padj < .05)

res_species_og <- res_species %>%
  as.data.frame %>%
  mutate(ortholog = rownames(.)) %>%
  filter(grepl("@", ortholog))
```

`r nrow(res_species_sig)` genes were significantly differentially abundant between wildmouse and 
labmouse, blocking for diet. `r table(res_species_sig$log2FoldChange > 0)[2]` were more abundant in wildmouse, while 
`r table(res_species_sig$log2FoldChange > 0)[1]` were more abundant in labmouse.

```{r write_tsvs}
write_tsv(res_sig, "outputs/diff_abund/mouse/mouse_diet_og_sig.tsv")
write_tsv(res_og, "outputs/diff_abund/mouse/mouse_diet_og_all.tsv")
write_tsv(res_species_sig, "outputs/diff_abund/mouse/mouse_species_og_sig.tsv")
write_tsv(res_species_og, "outputs/diff_abund/mouse/mouse_species_og_all.tsv")
```