---
title: "Observational Study KEGG Ortholog Results"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(echo = T, include = T, message = F, warning = F, cache = T)
```

## R Markdown

```{r libs}
library(tximport)
library(DESeq2)
library(clusterProfiler)
library(readr)
library(tidyr)
library(tibble)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(ggthemes)
library(ggbeeswarm)
library(UpSetR)
library(vegan)
library(kableExtra)
```


```{r tximport}
txi_kegg <- read_tsv("outputs/tx2gene/eggnog_tx2gene_kegg.tsv")
txi_kegg <- filter(txi_kegg, grepl("K", GENEID))
obs <- read_csv("inputs/tximport_samples/observational_samples.csv")
obs_counts <- tximport(files = obs$filename, type = "salmon", tx2gene = txi_kegg)
```

```{r export_counts}
obs_counts_raw <- obs_counts$counts
colnames(obs_counts_raw) <- obs$sample
obs_counts_raw <- as.data.frame(obs_counts_raw)
obs_counts_raw$ortholog <- rownames(obs_counts_raw)
write_tsv(obs_counts_raw, "outputs/ortholog_counts/obs_kegg_counts_raw.tsv")
```


```{r differential_abundance_status, include = F}
ddsTxi <- DESeqDataSetFromTximport(obs_counts,
                                   colData = obs,
                                   design = ~ status + diet_type + physiology) 
dds_status <- DESeq(ddsTxi)
# resultsNames(dds) # lists the coefficients
```

```{r norm_counts}
dds_kegg <- estimateSizeFactors(dds_status)
dds_kegg_counts <- counts(dds_kegg, normalized=TRUE)
colnames(dds_kegg_counts) <- colData(dds_kegg)$sample
dds_kegg_counts <- as.data.frame(dds_kegg_counts)
dds_kegg_counts$KO <- rownames(dds_kegg_counts)
write_tsv(dds_kegg_counts, "outputs/ortholog_counts/obs_kegg_counts_norm.tsv")
```

```{r pca_kegg}
vsd <- vst(dds_status)
pca_data <- plotPCA(vsd, intgroup = c("status", "species"), returnData=TRUE)
percentVar <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(PC1, PC2, shape=status, color=species)) + 
  theme_minimal() +
  geom_point(size=3) +
  geom_text_repel(aes(label = species)) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) +
  ggtitle("PCA plot of KEGG orthologs in observational study") +
  scale_color_tableau(palette = "Tableau 20")
```

```{r permanova}
vsd_assay <- t(assay(vsd))
rownames(vsd_assay) <- colData(vsd)$sample
veg_dist <- vegdist(vsd_assay, "bray")

obs <- obs[match(rownames(vsd_assay), obs$sample), ] # sort info by colnames
obs$species <- as.factor(obs$species)              # set factors for model
obs$status <- as.factor(obs$status)
obs$diet_type <- as.factor(obs$diet_type)
obs$physiology <- as.factor(obs$physiology)              # set factors for model
obs$dom_type <- as.factor(obs$dom_type)

perm <- adonis(veg_dist ~ dom_type + diet_type + physiology + status + species, 
               data = obs, 
               permutations = 10000)

write.table(perm$aov.tab, "outputs/permanova/observational_kegg_permanova.tsv", quote = F)

kable(perm$aov.tab) %>% kable_styling()
```

## Wild vs. domestic

```{r status_functions}
plot_status_sig <- function(KO, dds = dds_status, sig_df){
  KO_df <- plotCounts(dds, gene = KO, 
                      intgroup = c("status", "physiology", "diet_type"), 
                      normalized = TRUE, transform = TRUE, returnData=TRUE)
  
  ggplot(KO_df, aes(x = status, y = count, color = physiology, shape = diet_type)) + 
    geom_point(position=position_jitter(w = 0.1,h = 0)) +
    theme_bw() +
    ggtitle(KO) +
    theme(plot.title = element_text(hjust = 0.5)) +
    scale_y_log10() +
    labs(caption = paste0("log2 Fold Change = ", 
                          round(sig_df[sig_df$ortholog == KO, "log2FoldChange"],
                                digits = 3),
                          "; ", 
                          "p = ", 
                          round(sig_df[sig_df$ortholog == KO, "padj"], 
                                digits = 3)))
}
```
### Differential abundance

As we observed in the PCA plot, some wild/domestic species pairs had large differences
(e.g. bison and cow) while others had very small differences (e.g. wild sheep and 
domestic sheep). 

```{r res_diet}
res <- results(dds_status, name="status_wild_vs_domestic")
res <- res[order(res$pvalue), ]

res_sig <- res %>%
  as.data.frame %>%
  mutate(ortholog = rownames(.)) %>%
  filter(grepl("K", ortholog)) %>%         # keep only kegg orthologs
  filter(padj < .05) %>%                   # p value less than .05
  arrange(log2FoldChange)

res_kegg <- res %>%
  as.data.frame %>%
  mutate(ortholog = rownames(.)) %>%
  filter(grepl("K", ortholog))             # keep only kegg orthologs
```

When comparing wild vs. domesticated animals, there were `r nrow(res_sig)` 
differentially abundant KEGG 
orthologs of `r nrow(res_kegg)` annotated orthologs. Of these, 
`r table(res_sig$log2FoldChange < 0)[1]` were more abundant in wild
than domestic, while `r table(res_sig$log2FoldChange < 0)[2]` were less abundant 
in wild than domestic. 

However, these numbers only include orthologs with a p value less than .05. 
It is probably necessary to futher filter on log fold change or base mean of 
counts. For example, K06609,K08139,K08368 is significantly differentially abundant 
(p = `r round(res_kegg[res_kegg$ortholog == "ko:K06609,ko:K08139,ko:K08368", "padj"], digits = 3)`), 
but the log fold change is 
`r round(res_kegg[res_kegg$ortholog == "ko:K06609,ko:K08139,ko:K08368", "log2FoldChange"], digits = 3)`. It looks like this:

```{r plot_ortho_k02438}
plot_status_sig(KO = "ko:K06609,ko:K08139,ko:K08368", dds = dds_status, sig_df = res_kegg)
```


```{r res_sig_filt}
res_sig <- res_sig %>%
  filter(abs(log2FoldChange) > 3)
```  

This difference is not biologically meaningful. It may make sense to filter at to
an absolute log2 fold change greater than 3. At this level, there were 
`r nrow(res_sig)` differentially abundant KEGG orthologs of `r nrow(res_kegg)`
annotated orthologs. `r table(res_sig$log2FoldChange < 0)[1]` were more abundant 
in wild than domestic, while `r table(res_sig$log2FoldChange < 0)[2]` were less
abundantin wild than domestic. 

At this filtration level we see examples like K20370, which encodes a
phosphoenolpyruvate carboxykinase. This ortholog increased in abundance for 
some domestic animals. 

```{r plot_ortho, include = F, echo = F}
plot_status_sig(KO = "ko:K20370", dds = dds_status, sig_df = res_sig)
```

More examples of a differentially abundant KEGG orthologs are below. 

K01179,K01181,K01201,K07407 encodes a metabolic enzyme. In this experiment, it has a 
the largest absolute log2 Fold Change of the significantly differentially 
abundant KEGG orthologs (log2FC = `r min(res_sig$log2FoldChange)`, p < .001). 
We see that this means that this ortholog is less abundant in the wild species 
than the domestic species. 

In this case, we can see that this pattern is driven by the presence of the 
ortholog in 8 domestic samples present omnivoes and herbivores with hindgut 
physiology

```{r contrast_levels}
plot_status_sig(KO = "ko:K01179,ko:K01181,ko:K01201,ko:K07407", dds = dds_status, sig_df = res_sig)
```

Another example is K10953, RTX toxin RtxA associated with infection disease.
Again, we see that although this is more abundant in more domestic animals 
than in wild animals, this difference is driven by low abundance in some
wild sample. 

```{r contrast_level2}
plot_status_sig(KO = "ko:K10953", dds = dds_status, sig_df = res_sig)
```

### Enrichment 

Using KEGG ortholog pathway mappings, we can test for pathway enrichment 
in the differentially abundant orthologs. With the more stringent cut off of 
log2 fold change of 3, many pathways are enriched that are more abundant in wild
than domestic, but none are significantly enriched that are more abundant in domestic.

```{r enrichment_all}
## download ortholog to pathway table
pathways <- "http://rest.kegg.jp/link/pathway/ko"
download.file(url = pathways, destfile = "inputs/kegg_pathways.tsv")
pathways <- read_tsv("inputs/kegg_pathways.tsv", col_names = c("KO", "path"))
pathways <- pathways %>%
  mutate(KO = gsub("ko:", "", KO)) %>%
  filter(grepl(pattern = "map", path)) %>%
  #mutate(path = gsub("path:", "", path)) %>%
  dplyr::select(path, KO)

pathway_names <- "http://rest.kegg.jp/list/pathway"
download.file(url = pathway_names, destfile = "inputs/kegg_pathway_names.tsv")
pathway_names <- read_tsv("inputs/kegg_pathway_names.tsv", col_names = c("path", "name"))
```

```{r enrichment_diet}
res_ind <- res_sig %>% 
  filter(log2FoldChange > 0) %>%
  mutate(ortholog = gsub(".*,","", ortholog)) %>% # select last KO in annotation
  mutate(ortholog = gsub("ko:", "", ortholog))

enriched_ind <- enricher(gene = res_ind$ortholog, maxGSSize = 5000,
                         TERM2GENE = pathways, TERM2NAME = pathway_names)
clusterProfiler::dotplot(enriched_ind, font.size = 7, showCategory = nrow(enriched_ind@result)) +
  ggtitle("More abundant in wild")

res_rep <- res_sig %>% 
  filter(log2FoldChange < 0) %>%
  mutate(ortholog = gsub(".*,","", ortholog))%>%
  mutate(ortholog = gsub("ko:", "", ortholog))


enriched_rep <- enricher(gene = res_rep$ortholog, maxGSSize = 5000,
                         TERM2GENE = pathways, TERM2NAME = pathway_names)
clusterProfiler::dotplot(enriched_rep, font.size = 7, showCategory = nrow(enriched_rep@result)) +
  ggtitle("Less abundant in wild") 
```


```{r write_files}
## Diet contrast
write_tsv(res_sig, "outputs/diff_abund/obs_kegg_ortholog_sig.tsv")
write_tsv(res_kegg, "outputs/diff_abund/obs_kegg_ortholog_all.tsv")
write_tsv(enriched_ind@result, "outputs/diff_abund/obs_kegg_enrich_up.tsv")
write_tsv(enriched_rep@result, "outputs/diff_abund/obs_kegg_enrich_down.tsv")
```

```{r contrast_level_test, include = F, echo = F, eval = F}
K13274 <- plotCounts(dds_status, gene = "ko:K13274", 
                     intgroup = c("status", "physiology", "diet_type"), 
                     normalized = TRUE, transform = TRUE, returnData=TRUE)

ggplot(K13274, aes(x = status, y = count, color = physiology, shape = diet_type)) + 
  geom_point(position=position_jitter(w = 0.1,h = 0)) +
  theme_bw() +
  ggtitle("K13274") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_y_log10()
```

## Species-level comparisons

In the model we built above, we were able to include domestication status
(e.g. wild vs. domestic), physiology, and diet type. However, given that species
was fully contained within domestication status (e.g. all bison were wild, 
all cows were domestic, etc.), we could not include species in this model. 
However, instead of looking at differences in orthologs between wild and domestic
as a whole, we can look at differences between each species pair and then compare
which differentially abundant orthologs are shared between species pairs. 

This model is built only using the species variable. However, given that diet
and physiology are concordant between species pairs, and we are only testing
differences between pairs, other variables are described by the species variable.

```{r differential_abundance_species, include = F}
# ddsTxi_species <- DESeqDataSetFromTximport(obs_counts,
#                                           colData = obs,
#                                           design = ~ species) 
# dds_species <- DESeq(ddsTxi_species)
# 
# resultsNames(dds_species) # lists the coefficients
```


```{r function_analyze_pair}

calc_dds_species <- function(obs, wild, dom, tx2gene = txi_kegg){
  library(tximport)
  library(dplyr)
  library(DESeq2)
  
  obs <- obs %>% filter(species %in% c(wild, dom))
  counts <- tximport(files = obs$filename, type = "salmon", tx2gene = tx2gene)
  ddsTxi <- DESeqDataSetFromTximport(counts,
                                     colData = obs,
                                     design = ~ species) 
  dds <- DESeq(ddsTxi)
  return(dds)
}

calc_species_sig <- function(dds, wild, dom){
  library(dplyr)
  library(DESeq2)
  
  species_res <- results(dds, name = resultsNames(dds)[2])
  species_sig <- species_res %>%
    as.data.frame %>%
    mutate(ortholog = rownames(.)) %>%
    filter(grepl("K", ortholog)) %>%       # keep only kegg orthologs
    filter(padj < .05)                     # p value less than .05
  
  species_ind <- filter(species_sig, log2FoldChange > 0)
  species_rep <- filter(species_sig, log2FoldChange < 0)
  
  return(list(significant_orthologs = species_sig,
              more_abundant = species_ind,
              less_abundant = species_rep))
}

species_enrich <- function(species_sig, 
                           TERM2GENE = pathways, TERM2NAME = pathway_names, 
                           wild, dom){
  library(DESeq2)
  library(clusterProfiler)
  library(readr)
  
  species_ind <- species_sig %>% 
    filter(log2FoldChange > 0) %>%
    mutate(ortholog = gsub(".*,","", ortholog)) %>%
    mutate(ortholog = gsub("ko:", "", ortholog))
  
  species_enriched_ind <- enricher(gene = species_ind$ortholog, maxGSSize = 5000,
                               TERM2GENE = TERM2GENE, TERM2NAME = TERM2NAME)
  write_tsv(species_enriched_ind@result, 
            paste0("outputs/diff_abund/observational_species/", wild, "_vs_", dom, "_more.tsv"))
  plot1 <- clusterProfiler::dotplot(species_enriched_ind, font.size = 7,
                                    showCategory = nrow(species_enriched_ind@result)) +
    ggtitle("More abundant in wild")

  species_rep <- species_sig %>%
    filter(log2FoldChange < 0) %>%
    mutate(ortholog = gsub(".*,","", ortholog)) %>%
    mutate(ortholog = gsub("ko:", "", ortholog))

  species_enriched_rep <- enricher(gene = species_rep$ortholog, maxGSSize = 5000,
                                   TERM2GENE = TERM2GENE, TERM2NAME = TERM2NAME)
  write_tsv(species_enriched_rep@result, 
            paste0("outputs/diff_abund/observational_species/", wild, "_vs_", dom, "_less.tsv"))
  plot2 <- clusterProfiler::dotplot(species_enriched_rep, font.size = 7,
                                    showCategory = nrow(species_enriched_rep@result)) + 
    ggtitle("Less abundant in wild")
  
  return(list(more_abund_plot = plot1,
              less_abund_plot = plot2))
}

plot_species_sig <- function(KO, dds, species_sig, wild, dom){
  KO_df <- plotCounts(dds, gene = KO, 
                       intgroup = "species", 
                       normalized = TRUE, transform = TRUE, returnData=TRUE)

  ggplot(KO_df %>% filter(species %in% c(wild, dom)), 
         aes(x = species, y = count)) + 
    geom_point(position=position_jitter(w = 0.2,h = 0)) +
    theme_bw() +
    ggtitle(KO) +
    theme(plot.title = element_text(hjust = 0.5)) +
    scale_y_log10() +
    labs(caption = paste0("log2 Fold Change = ", 
                          round(species_sig[species_sig$ortholog == KO, "log2FoldChange"],
                                digits = 3),
                          "; ", 
                          "p = ", 
                          round(species_sig[species_sig$ortholog == KO, "padj"], 
                                digits = 3)))
}

```



### Boar/pig


```{r pig_analysis}
pig_dds <- calc_dds_species(obs = obs, wild = "boar", dom = "pig")
pig_sig <- calc_species_sig(dds = pig_dds, wild = "boar", dom = "pig")
```

In boar vs. pig, `r nrow(pig_sig$significant_orthologs)` were differentially
abundant. `r nrow(pig_sig$more_abundant)` were more abundant in boar,
while `r nrow(pig_sig$less_abundant)` were more abundant in pig.

We can also see from a plot of counts from a KEGG ortholog with a low log2 
fold change. Unlike in the wild vs. domestic contrast, we see that small log2
fold change reflects real differences. Therefore, we filter at a p value of .05, 
allowing significant differences of any log2 fold change to be considered 
significant.


```{r pig_enrich}
plot_species_sig(KO = "ko:K03558", dds = pig_dds, 
                 species_sig = pig_sig$significant_orthologs,
                 wild = "boar", dom = "pig")
```

We also see that there are many enriched pathways in the differences between boar
and pig.

```{r}
pig_enrich <- species_enrich(species_sig = pig_sig$significant_orthologs, 
                             wild = "boar", dom = "pig")
pig_enrich[[1]]
pig_enrich[[2]]
```

### Wildmouse/labmouse

```{r mouse_analysis}
mouse_dds <- calc_dds_species(obs = obs, wild = "wildmouse", dom = "labmouse")
mouse_sig <- calc_species_sig(dds = mouse_dds, wild = "wildmouse", dom = "labmouse")
```

In wildmouse vs. labmouse, 
`r nrow(mouse_sig$significant_orthologs)` were differentially abundant. 
`r nrow(mouse_sig$more_abundant)` were more abundant in wildmouse, while 
`r nrow(mouse_sig$less_abundant)` were more abundant in labmouse.


```{r mouse_enrich}
plot_species_sig(KO = "ko:K01593,ko:K01594", dds = mouse_dds, 
                 species_sig = mouse_sig$significant_orthologs,
                 wild = "wildmouse", dom = "labmouse")
```

We also see that there are many enriched pathways in the differences between wildmouse
and labmouse.

```{r}
mouse_enrich <- species_enrich(species_sig = mouse_sig$significant_orthologs, 
                               wild = "wildmouse", dom = "labmouse")
mouse_enrich[[1]]
mouse_enrich[[2]]
```

## Bison and cow

```{r cow_analysis}
cow_dds <- calc_dds_species(obs = obs, wild = "bison", dom = "cow")
cow_sig <- calc_species_sig(dds = cow_dds, wild = "bison", dom = "cow")
```

In bison vs. cow, 
`r nrow(cow_sig$significant_orthologs)` were differentially abundant. 
`r nrow(cow_sig$more_abundant)` were more abundant in bison, while 
`r nrow(cow_sig$less_abundant)` were more abundant in cow.

K01179,K03768,K08738 encodes an endoglucanase and has the largest absolute
log2 Fold Change between bison and cow. 

```{r cow_enrich}
plot_species_sig(KO = "ko:K01179,ko:K03768,ko:K08738", dds = cow_dds, 
                 species_sig = cow_sig$significant_orthologs,
                 wild = "bison", dom = "cow")
```

We also see that there are many enriched pathways in the differences between bison
and cow.

```{r}
cow_enrich <- species_enrich(species_sig = cow_sig$significant_orthologs, 
                               wild = "bison", dom = "cow")
cow_enrich[[1]]
cow_enrich[[2]]
```

## Wolf and dog

```{r dog_analysis}
dog_dds <- calc_dds_species(obs = obs, wild = "wolf", dom = "dog")
dog_sig <- calc_species_sig(dds = dog_dds, wild = "wolf", dom = "dog")
```

In wolf vs. dog, 
`r nrow(dog_sig$significant_orthologs)` were differentially abundant. 
`r nrow(dog_sig$more_abundant)` were more abundant in wolf, while 
`r nrow(dog_sig$less_abundant)` were more abundant in dog.

K10175 encodes a TBX1, and has the largest absolute log2 Fold Change
between dog and wolf.

```{r dog_enrich}
plot_species_sig(KO = "ko:K10175", dds = dog_dds, 
                 species_sig = dog_sig$significant_orthologs,
                 wild = "wolf", dom = "dog")
```

We also see that there are many enriched pathways in the differences between wolf
and dog.

```{r}
dog_enrich <- species_enrich(species_sig = dog_sig$significant_orthologs, 
                               wild = "wolf", dom = "dog")
dog_enrich[[1]]
dog_enrich[[2]]
```

## Sheep

```{r sheep_analysis}
sheep_dds <- calc_dds_species(obs = obs, wild = "sheep_wild", dom = "sheep_dom")
sheep_sig <- calc_species_sig(dds = sheep_dds, wild = "sheep_wild", dom = "sheep_dom")
```

In sheep_wild vs. sheep_dom, 
`r nrow(sheep_sig$significant_orthologs)` were differentially abundant. 
`r nrow(sheep_sig$more_abundant)` were more abundant in sheep_wild, while 
`r nrow(sheep_sig$less_abundant)` were more abundant in sheep_dom. 

 	
K02850 encodes heptose II phosphotransferase, and has the largest
positive log2 Fold Change between domesticated and wild sheep.
	
```{r sheep_enrich}
plot_species_sig(KO = "ko:K02850", dds = sheep_dds, 
                 species_sig = sheep_sig$significant_orthologs,
                 wild = "sheep_wild", dom = "sheep_dom")
```

We also see that there are many enriched pathways in the differences 
between sheep_wild and sheep_dom.

```{r}
sheep_enrich <- species_enrich(species_sig = sheep_sig$significant_orthologs, 
                               wild = "sheep_wild", dom = "sheep_dom")
sheep_enrich[[1]]
sheep_enrich[[2]]
```


## Guineapig

```{r guineapig_analysis}
guineapig_dds <- calc_dds_species(obs = obs, wild = "guineapig_wild", dom = "guineapig_lab")
guineapig_sig <- calc_species_sig(dds = guineapig_dds, wild = "guineapig_wild", dom = "guineapig_lab")
```

In guineapig_wild vs. guineapig_lab, 
`r nrow(guineapig_sig$significant_orthologs)` were differentially abundant. 
`r nrow(guineapig_sig$more_abundant)` were more abundant in guineapig_wild, while 
`r nrow(guineapig_sig$less_abundant)` were more abundant in guineapig_lab.

K10297,K15923 has the largest absolute log2 Fold change between lab and wild 
guineapig. It encodes an F box protein or alpha-L-fucosidase 2.

```{r guineapig_enrich}
plot_species_sig(KO = "ko:K10297,ko:K15923", dds = guineapig_dds, 
                 species_sig = guineapig_sig$significant_orthologs,
                 wild = "guineapig_wild", dom = "guineapig_lab")
```

We also see that there are many enriched pathways in the differences between
guineapig_wild and guineapig_lab.

```{r}
guineapig_enrich <- species_enrich(species_sig = guineapig_sig$significant_orthologs, 
                               wild = "guineapig_wild", dom = "guineapig_lab")
guineapig_enrich[[1]]
guineapig_enrich[[2]]
```


## Rat

```{r rat_analysis}
rat_dds <- calc_dds_species(obs = obs, wild = "rat_wild", dom = "rat_lab")
rat_sig <- calc_species_sig(dds = rat_dds, wild = "rat_wild", dom = "rat_lab")
```

In rat_wild vs. rat_lab, 
`r nrow(rat_sig$significant_orthologs)` were differentially abundant. 
`r nrow(rat_sig$more_abundant)` were more abundant in rat_wild, while 
`r nrow(rat_sig$less_abundant)` were more abundant in rat_lab.

K12293,K20345 has the largest absolute log2 fold change between lab rat and wild
rat. It encodes a competence factor transport accessory protein ComB or membrane 
fusion protein, peptide pheromone/bacteriocin exporter.
```{r rat_enrich}
plot_species_sig(KO = "ko:K12293,ko:K20345", dds = rat_dds, 
                 species_sig = rat_sig$significant_orthologs,
                 wild = "rat_wild", dom = "rat_lab")
```

We also see that there are many enriched pathways in the differences between rat_wild
and rat_lab.

```{r}
rat_enrich <- species_enrich(species_sig = rat_sig$significant_orthologs, 
                               wild = "rat_wild", dom = "rat_lab")
rat_enrich[[1]]
rat_enrich[[2]]
```


## Rabbit


```{r rabbit_analysis}
rabbit_dds <- calc_dds_species(obs = obs, wild = "rabbit_wild", dom = "rabbit_dom")
rabbit_sig <- calc_species_sig(dds = rabbit_dds, wild = "rabbit_wild", dom = "rabbit_dom")
```

In rabbit_wild vs. rabbit_dom, 
`r nrow(rabbit_sig$significant_orthologs)` were differentially abundant. 
`r nrow(rabbit_sig$more_abundant)` were more abundant in rabbit_wild, while 
`r nrow(rabbit_sig$less_abundant)` were more abundant in rabbit_dom.

K00261 has the largest absoluate log2 fold change between wild and 
domesticated rabbit. It encodes a glutamate dehydrogenase (NAD(P)+).

```{r rabbit_enrich}
plot_species_sig(KO = "ko:K00261", dds = rabbit_dds, 
                 species_sig = rabbit_sig$significant_orthologs,
                 wild = "rabbit_wild", dom = "rabbit_dom")
```

We also see that there are many enriched pathways in the differences between rabbit_wild
and rabbit_dom.

```{r}
rabbit_enrich <- species_enrich(species_sig = rabbit_sig$significant_orthologs, 
                               wild = "rabbit_wild", dom = "rabbit_dom")
rabbit_enrich[[1]]
rabbit_enrich[[2]]
```

### Comparing between species pairs

We can compare the number of differentially abundant orthologs in common 
between species pairs.

#### More abundant in wild

We see that for orthologs that are more abundant in wild, the majority 
are significantly different in a single species pair. However, this is a large
overlap between guineapig and rat, and between rat, mouse, and guineapig, and bison/cow and boar/pig.

```{r upset_more_abund, fig.width = 12}

# generate a list for each species for each "more abundant in wild" ortholog.

list_more_abund_in_wild <- list(
  bison_vs_cow = cow_sig$more_abundant$ortholog,
  wolf_vs_dog  = dog_sig$more_abundant$ortholog,
  boar_vs_pig  = pig_sig$more_abundant$ortholog,
  mouse        = mouse_sig$more_abundant$ortholog,
  rat          = rat_sig$more_abundant$ortholog,
  guineapig    = guineapig_sig$more_abundant$ortholog,
  sheep        = sheep_sig$more_abundant$ortholog,
  rabbit       = rabbit_sig$more_abundant$ortholog)

upset(fromList(list_more_abund_in_wild), order.by = "degree",
      nsets = 8, nintersects = 200)
```

Comparitively, we see more overlap in orthologs that are more abundant in 
domesticated animals. Although these numbers are smaller, there are fewer
orthologs that were more abundant in the domesticated animals than in the wild
animals. We also see that the number of sets that overlap increases, with as
many as six sets sharing the same orthologs.

```{r upset_less_abund, fig.width = 12}

list_less_abund_in_wild <- list(
  bison_vs_cow = cow_sig$less_abundant$ortholog,
  wolf_vs_dog  = dog_sig$less_abundant$ortholog,
  boar_vs_pig  = pig_sig$less_abundant$ortholog,
  mouse        = mouse_sig$less_abundant$ortholog,
  rat          = rat_sig$less_abundant$ortholog,
  guineapig    = guineapig_sig$less_abundant$ortholog,
  sheep        = sheep_sig$less_abundant$ortholog,
  rabbit       = rabbit_sig$less_abundant$ortholog)

upset(fromList(list_less_abund_in_wild), order.by = "degree",
      nsets = 8, nintersects = 200)
```

## KEGG pathway analysis

```{r read_kegg_counts_to_pathways}
# create a tximport file for pathway info using the eggnog file
eggnog <- read_tsv("outputs/atlas_out/Genecatalog/annotations/eggNog.tsv.gz")
txi_kegg_pathway <- eggnog %>%
  select(TXNAME = Query, GENEID=KEGG_Pathway) %>%
  filter(!is.na(GENEID)) %>%
  mutate(GENEID = gsub(",.*", "", GENEID)) %>% # select the first pathway :( 
  mutate(GENEID = gsub("ko", "map", GENEID))
obs <- read_csv("inputs/tximport_samples/observational_samples.csv")
obs_pathways <- tximport(files = obs$filename, type = "salmon", tx2gene = txi_kegg_pathway)
```


```{r}
# re-read pathway names
pathway_names <- read_tsv("inputs/kegg_pathway_names.tsv", col_names = c("path", "name")) %>%
  mutate(path = gsub("path:", "", path))
```

```{r ko_sums}
# calculate the number of kegg ortholog counts observed in each library
# this will become the denominator for 
obs_counts <- read_tsv("outputs/ortholog_counts/obs_kegg_counts_raw.tsv")
ko_sums <- obs_counts %>%
  select(-ortholog) %>%
  mutate(across(everything(), ~replace_na(.x, 0))) %>%
  summarise(across(everything(), list(sum))) %>%
  rename_all(~gsub("_1$", "", .)) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  mutate(across(where(is.numeric), ceiling)) %>%
  select(sample, num_reads = V1)
```

```{r}
# filter and prune obs metadata to the columns we still need, and join with read 
# to be used as denominator for corncob
obs <- obs %>%
  select(sample, species, dom_status = status, diet_type, physiology, dom_type) %>%
  left_join(ko_sums, by = "sample")
```

```{r }
# grab pathway abundances
obs_pathways <- obs_pathways$counts
colnames(obs_pathways) <- obs$sample
obs_pathways <- as.data.frame(obs_pathways) %>%
  mutate(across(where(is.numeric), ceiling))
```

```{r prep_for_corncob}
# transpose counts
obs_pathways_t <- obs_pathways %>% 
  t() %>% 
  as.data.frame() %>%
  rownames_to_column("sample")

# join with metadata
df <- left_join(obs, obs_pathways_t, by = "sample")
```

```{r run_corncob}
fit_corncob <- function(col_num) {
  print(col_num)
  # record protein being queried
  pathway_name <- df %>%
    select(all_of(col_num)) %>%
    colnames()
  # run corncob with LRT
  try({
    corncob_out <- df %>%
      select(species, dom_status, diet_type, physiology, dom_type, num_reads, all_of(col_num)) %>%
      dplyr::rename(ww = 7) %>%
      corncob::bbdml(formula = cbind(ww, num_reads - ww) ~ dom_status + diet_type + physiology, 
                     formula_null = cbind(ww, num_reads - ww) ~ diet_type + physiology,
                     phi.formula = ~ 1,
                     phi.formula_null = ~ 1,
                     data = .,
                     test = "LRT", 
                     boot = FALSE) %>%
      summary()
    corncob_coeff <- data.frame(mu = rownames(corncob_out$coefficients), 
                                estimate = corncob_out$coefficients[ , 'Estimate'],
                                standard_error = corncob_out$coefficients[ , 'Std. Error'],
                                t_value = corncob_out$coefficients[ , "t value"],
                                p_value = corncob_out$coefficients[ , "Pr(>|t|)"],
                                pathway = pathway_name)
    return(corncob_coeff)
  }) # end try
}

# run function on all AAs
all_ccs <- sapply(8:ncol(df), fit_corncob, simplify=F)
all_ccs <- do.call(rbind, all_ccs)

# perfom pvalue adustment
sig2 <- all_ccs %>%
  filter(mu %in% c("mu.dom_statuswild")) %>%
  group_by(mu) %>%
  mutate(fdr = p.adjust(p_value, method = "fdr")) %>%
  filter(fdr < .05) %>%
  left_join(pathway_names, by = c("pathway" = "path"))
View(sig2)
```


### See how results change if we pick the second pathway
It does ok, but seems less specific (e.g. "carbon metabolism" instead of "starch and sucrose metabolism")
The first method recovers 26 significant pathways, the second 51. 
They overlap by 10 pathways.
```{r read_kegg_counts_to_pathways, eval = F}
# create a tximport file for pathway info using the eggnog file
eggnog <- read_tsv("outputs/atlas_out/Genecatalog/annotations/eggNog.tsv.gz")
txi_kegg_pathway <- eggnog %>%
  select(TXNAME = Query, GENEID=KEGG_Pathway) %>%
  filter(!is.na(GENEID)) %>%
  mutate(GENEID = basename(gsub(",", "\\/", GENEID))) # select the last pathway :( 

obs <- read_csv("inputs/tximport_samples/observational_samples.csv")
obs_pathways <- tximport(files = obs$filename, type = "salmon", tx2gene = txi_kegg_pathway)

# grab pathway abundances
obs_pathways <- obs_pathways$counts
colnames(obs_pathways) <- obs$sample
obs_pathways <- as.data.frame(obs_pathways) %>%
  mutate(across(where(is.numeric), ceiling))

# transpose counts
obs_pathways_t <- obs_pathways %>% 
  t() %>% 
  as.data.frame() %>%
  rownames_to_column("sample")

obs <- obs %>%
  select(sample, species, dom_status = status, diet_type, physiology, dom_type) %>%
  left_join(ko_sums, by = "sample")

# join with metadata
df <- left_join(obs, obs_pathways_t, by = "sample")

# run function on all AAs
all_ccs <- sapply(8:ncol(df), fit_corncob, simplify=F)
all_ccs <- do.call(rbind, all_ccs)
View(all_ccs)

# perfom pvalue adustment
sig <- all_ccs %>%
  filter(mu %in% c("mu.dom_statuswild")) %>%
  group_by(mu) %>%
  mutate(fdr = p.adjust(p_value, method = "fdr")) %>%
  filter(fdr < .05) %>%
  left_join(pathway_names, by = c("pathway" = "path"))

```
